---
author: "Zhenya Ratushko"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}


## Assignment 7

#### Due Saturday, October 28, 11:59 PM CT

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw07/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw07/hw07.Rmd
  - COURSE/scripts/viridis.R
  - COURSE/scripts/ggprob.R

### Aims

- Practice probability distributions

### Problems

######  1.
A discrete random variable $X$ has possible values 0, 1, 2, 3, 4, 5, 6, 7, 8 with the following partial probability distribution.
The missing probabilities $\prob(X=7)$ and $\prob(X=8)$ are equal to each other.

```{r, echo = FALSE}
set.seed(20221021)
x = 0:8
p = rnorm(7,5,2.1)
p = sort(round(p / sum(p) * 0.84, 2))
p1 = tibble(x = x, p = c(p, rep(NA,2)))

p1_tab = p1 %>% 
  mutate(x = str_c(x)) %>% 
  pivot_wider(names_from = x, values_from = p) %>% 
  mutate(x = "P(X=x)", .before = `0`) 

p1_tab %>% 
  kable() %>% 
  kable_styling(position = "center", full_width = FALSE)
```

- Put the values `x = 0:8` and the tabled probabilities `p` into a data frame (use `tibble()`).

```{r}
q1 = tibble(x = 0:8, p = c(0.04, 0.06, 0.11, 0.12, 0.14, 0.17, 0.2, NA, NA))
q1
```

- Write code to calculate the missing values for $\prob(X=7) = \prob(X = 8)$ and replace the missing values with these calculated values into your data frame.

```{r}
partial_sum = q1 %>% 
  filter(x != 7, x != 8) %>% 
  summarize(sum_p = sum(p)) %>% 
  pull(sum_p)

p7and8 = (1 - partial_sum)/2

q1 = q1 %>% 
  mutate(p = case_when(
    !is.na(p) ~ p,
    TRUE ~ p7and8))

q1
```

- Verify that the sum of all probabilities is 1.

```{r}
q1 %>%
  summarize(total = sum(p))
```



###### 2.
For the probability distribution from the previous problem:
  
- Calculate the mean, variance, and standard deviation.

```{r}
q2 = q1 %>% 
  summarize(mu = sum(x*p), sigma2 = sum((x-mu)^2*p), sigma = sqrt(sigma2))

q2
```




###### 3.
For the probability distribution from Problem 1:
  
- Draw a graph of the probability distribution with line segments at each possible value.
- Add to the graph a dashed red line at the mean and dotted black lines that are one and two standard deviations above and below the mean.
  
  
```{r}
ggplot(q1, aes(x = x, y = p)) +
  geom_segment(aes(xend = x, yend = 0), color = "blue", size = 2) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 4.45, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 4.45 + c(-1,1) * 2.123087, color = "black", linetype = "dotted") +
  geom_vline(xintercept = 4.45 + c(-2,2) * 2.123087, color = "black", linetype = "dotted") +
  scale_x_continuous(breaks = 0:8) +
  ylab("P(X=x)") +
  ggtitle("Distribution of X")
```
  



###### 4.
For the probability distribution from Problem 1:
  
- Add to the `tibble` for the probability distribution for the random variable $X$ a column named `cdf` which contains the value $F(x) = \prob(X \le x)$.
    - Note that for this example, $F(0) = \prob(X = 0) = 0.04$ and $F(1) = \prob(X = 0) + \prob(X = 1) = 0.10$.
- Print the entire modified data frame.

```{r}
q4 = q1 %>%
  mutate(cdf = cumsum(p))

print(q4, n = Inf)
```


The function $F(x)$ is known as the *cumulative distribution function*. In this example, $F(x) = 0$ for all $x < 0$, $F(x) = 1$ for all $x > 8$, and $0 < F(x) < 1$ if $0 < x < 8$. The function `cumsum()` may be very useful when calculating $F(x)$.
  


###### 5.
Using the probability distribution from the previous problem:
   
- Draw a graph to display the cumulative distribution function $F(x)$ with a step function.
    - For the purposes of the graph only, it may be helpful to add rows with $x=-1$ and $x=9$.
    - The **ggplot2** function `geom_step()` will be very useful to make this plot. See the help page or search the internet for an example on how to use it.
    
    
```{r}
ggplot(q4, aes(x = x, y = cdf)) +
  geom_step() +
  scale_y_continuous(breaks = seq(0, 1, 0.2))
```



###### 6.
The $p$ quantile of a probability distribution is any location $a$ where $\prob(X \le a) \ge p$ and $\prob(X \ge a) \ge 1-p$. For a continuous random variable with a density function, ther is typically a one-to-one correspondence between the quantile $a$ and the probability $p$. In contrast, for a discrete distribution such as the one featured in this assignment, the same value $a$ may be the quantile for many different $p$ and a few key values of $p$ have multiple possible quantile values $a$. The quantile function is essentially the inverse function $F^{-1}(p)$, which maps using the graph of $F$ from $p$ on the y-axis to the quantile $a$ on the x-axis.
  
- By examining the graph of the CDF in the previous problem, identify the 0.2, 0.3, 0.4, and 0.5 quantiles of the probability distribution of $X$. (You can do this without code, but a pat on the back if you can write code to find the values.)

> Regarding the probability distribution of X, the 0.2 quantile is 2, the 0.3 quantile is 3, the 0.4 quantile is 4, and the 0.5 quantile is 5; this is because the horizontal lines created for each quantile value cross the vertical lines corresponding to each aforementioned respective value of the function (see graph).

- Explain why there are multiple values $a$ which are 0.1 quantiles of the distribution and specify all such values.

> There are multiple values of a which are 0.1 quantiles of the distribution because the horizontal line for 0.1 directly overlays the line of the function stretching from the values of 1 and 2, meaning that the 0.1 quantile includes any value between that range (see graph).

```{r}
ggplot(q4, aes(x = x, y = cdf)) +
  geom_step() +
  scale_y_continuous(breaks = seq(0, 1, 0.2)) +
  geom_hline(yintercept = 0.1, color = "brown") +
  geom_hline(yintercept = 0.2, color = "purple") +
  geom_hline(yintercept = 0.3, color = "magenta") +
  geom_hline(yintercept = 0.4, color = "skyblue") +
  geom_hline(yintercept = 0.5, color = "navy") 
```



###### 7.
The code below draws a graph of a discrete distribution called the *binomial distribution* with parameters $n = 5$ and $p = 0.4$. The median of the distribution is the 0.5 quantile. 
  
```{r}
gbinom(5, 0.4)
qbinom(0.5, 5, 0.4)
```
  
- Based on the graph, what value of $x$ does the median have? Compare to the value of `qbinom(0.5, 5, 0.4)`.

> Based on the graph, the median value appears to be 2, because that value is the most-occurring one in the binomial distribution, making it plausible that it would also be the median. This is proven by the result that is outputted by `qbinom(0.5, 5, 0.4)`, which is also 2.

###### 8.
The code below draws a graph of a continuous normal distribution with parameters $\mu = 50$ and $\sigma = 10$ with the area to the left of 60 shaded.
  
```{r}
gnorm(50, 10) +
  geom_norm_fill(mu = 50, sigma = 10, b = 60) +
  scale_x_continuous(breaks = seq(30, 70, 10))
```
  
The value of `pnorm(60, 50, 10)` is the area to the left of 60 under this density curve.

```{r}
pnorm(60, 50, 10)
```

- Is the 0.80 quantile of this distribution smaller than, equal to, or greater than 60? Briefly explain.

> The 0.80 quantile of this distribution is smaller than 60, because the value of `pnorm(60, 50, 10)` is around 0.84, making it 0.04 larger than 0.80; the value of a call to `pnorm` would need to be exactly 0.80 to be equal to the 0.80 quantile, a value smaller than 0.80 to be smaller than the 0.80 quantile, and a value bigger than 0.80 to be bigger than the 0.80 quantile (as shown here).
    
 